---
title: "Covid-19 Data"
author: "Taylor Preusch"
date: "9/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tinytex)
```

## R Markdown

The following document shows the analysis of vaccines and stay at home orders on covid 19 deaths and cases. I have chosen this analysis for two reasons. The first reason as both are highly topical datasets and I am personally curious about the impact on both and the spread of covid. I have biases about each that I would like to address with this analysis. The second reason is that the John Hopkins data is daily and these two datasets are also calculated daily. This means that a meaningful correlation could help influence policy around both.

The first step I took was getting data from reproducible sources. I was able to get good data from both government agencies and news sources. 

The second step was to clean the data, pivot, and group it by the same grain. All analysis is daily by state. 

I then manufactured features for the model.
I normalized the data against the total population where it made sense like vaccinations and population mobility.

I created new features to predict like percent change and number change in both cases and deaths. Some very disturbing visualizations can be made by using the total cases and deaths. For example in both cases my graphs show that covid cases have increase with vaccinations. This is 100% true but could be extremely misleading if vaccinations were presumed to cause covid. 

I split my data into a training and validation set to predict covid cases and deaths in the 4 corner states: Colorado, Arizona, New Mexico, and Utah. 

I then visualized the prediction against our primary variables mobility and vaccinations.

Conclusion: So far in 2021 of the external variables, vaccines have had the largest impact on the transmission of COVID. Across all predicted features it was statistically significant.

The percent of people staying at home vs mobile hasn't changed dramatically throughout the year. This data only proves to be statistically significant in predicting cases but does not help predict deaths.

Next Action: There are many more cuts of the mobility data including how far somebody has traveled. I would like to run analysis on this to see if states with more long distance travel have a higher prevelance of covid or if local travel is the major determinant. 


```{r data creation}

url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- c("time_series_covid19_confirmed_global.csv",
                "time_series_covid19_deaths_global.csv",
                "time_series_covid19_confirmed_US.csv",
                "time_series_covid19_deaths_US.csv")
urls <- str_c(url_in,file_names)
urls
```


```{r data download}

raw_global_cases <- read_csv(urls[1])
raw_global_deaths <- read_csv(urls[2])
raw_us_cases <- read_csv(urls[3])
raw_us_deaths <- read_csv(urls[4])

```


```{r data creation vaccines and trips}

current_usa <- read_csv("https://raw.githubusercontent.com/BloombergGraphics/covid-vaccine-tracker-data/master/data/current-usa.csv")
usa_vaccine_time_series_by_state <- read_csv("https://raw.githubusercontent.com/BloombergGraphics/covid-vaccine-tracker-data/master/data/historical-usa-doses-administered.csv")
Trips_by_Distance <- read_csv("https://data.bts.gov/api/views/w96p-f2qv/rows.csv?accessType=DOWNLOAD")

```
```{r trip data creation}

by_level <- Trips_by_Distance %>% group_by(Level)

by_level_mean <- by_level %>% summarise(
  pop_not_home = mean(`Population Not Staying at Home`),
  pop_at_home = mean(`Population Staying at Home`)
)

by_state <- filter(Trips_by_Distance, Level == "State") %>% group_by(`State Postal Code`,`Date`)

by_state_mean <- by_state %>% summarise(
  pop_not_home = mean(`Population Not Staying at Home`),
  pop_at_home = mean(`Population Staying at Home`)
) %>% rename("date" = "Date") %>% rename("stateAbbrev" = "State Postal Code")


```

```{r vaccine data creation}

usa_vaccine_time_series_by_state <- usa_vaccine_time_series_by_state %>% rename("stateAbbrev" = "id") %>% rename("vaccinations" = "value")

```

```{r pivots cleaning}
global_cases <- select(raw_global_cases, -1, -3, -4)
us_cases <- select(raw_us_cases, -1:-6,-8,-9, -10, -11)

global_deaths <- select(raw_global_deaths, -1, -3, -4)
us_deaths <- select(raw_us_deaths, -1:-6,-8,-9, -10, -11)

```

```{r pivots}

global_cases_pivot <- global_cases %>% pivot_longer(!"Country/Region", names_to = "date", values_to = "cases") %>% mutate(date = mdy(date))

global_deaths_pivot <- global_deaths %>% pivot_longer(!"Country/Region", names_to = "date", values_to = "cases") %>% mutate(date = mdy(date)) %>% rename("Country_Region" = "Country/Region")




us_cases_pivot <- us_cases %>% pivot_longer(!"Province_State", names_to = "date", values_to = "cases") %>% mutate(date = mdy(date))

us_cases_group <- us_cases_pivot %>% group_by(`Province_State`,`date`) %>% summarise(
  total_cases = sum(`cases`)
) 

us_cases_group$stateAbbrev <- state.abb[match(us_cases_group$Province_State,state.name)]

us_deaths_pivot <- us_deaths %>% pivot_longer(!c("Province_State", "Population"), names_to = "date", values_to = "cases") %>% mutate(date = mdy(date))

us_deaths_group <- us_deaths_pivot %>% group_by(`Province_State`,`date`) %>% summarise(
  total_population = sum(`Population`),
  total_deaths = sum(`cases`)
) 

us_deaths_group$stateAbbrev <- state.abb[match(us_deaths_group$Province_State,state.name)]


```

```{r population corrected data}

us_regression_data <- us_deaths_group %>% inner_join(by_state_mean,by=c("date","stateAbbrev"))
us_regression_data <- us_regression_data %>% inner_join(us_cases_group,by=c("date","stateAbbrev"))
us_regression_data <- us_regression_data %>% inner_join(usa_vaccine_time_series_by_state,by=c("date","stateAbbrev"))

us_regression_data$vaccine_rate <- us_regression_data$vaccinations / us_regression_data$total_population
us_regression_data$home_rate <- us_regression_data$pop_at_home / us_regression_data$total_population 
us_regression_data$not_home_rate <- us_regression_data$pop_not_home / us_regression_data$total_population
us_regression_data <- us_regression_data %>% 
     mutate(pct_change_cases = ((total_cases/lag(total_cases) - 1) * 100))

us_regression_data <- us_regression_data %>% 
     mutate(pct_change_death = ((total_deaths/lag(total_deaths) - 1) * 100))

us_regression_data <- us_regression_data %>% 
     mutate(number_change_death = (total_deaths - lag(total_deaths)))

us_regression_data <- us_regression_data %>% 
     mutate(number_change_cases = (total_cases - lag(total_cases)))

us_regression_data <-  filter(us_regression_data, date >= "2021-01-01")
us_regression_data <-  filter(us_regression_data, stateAbbrev == "CO" | stateAbbrev == "UT" | stateAbbrev == "AZ"| stateAbbrev == "NM")
us_regression_data_selected <- select(us_regression_data, -1,-2,-3,-6,-7, -8, -10)



```

## Stay at Home vs Leave the Home vs Vaccinations


```{r model plot}

us_regression_data %>% ggplot() + geom_point(aes(x = date, y = home_rate), color = "red")+
  geom_point(aes(x = date, y = not_home_rate), color = "blue")+geom_point(aes(x = date, y = vaccine_rate), color = "purple")

```

## Training Data Creation

```{r sampling }
set.seed(222)
 
sample_size = round(nrow(us_regression_data_selected)*.70) # setting what is 70%
index <- sample(seq_len(nrow(us_regression_data_selected)), size = sample_size)
 
train <- us_regression_data_selected[index, ]
test <- us_regression_data_selected[-index, ]

```



## Model Creation


```{r data model total cases}
mod <- lm(total_cases ~ vaccine_rate+home_rate+not_home_rate, data = train)

summary(mod)
```
```{r data model number change in cases}
mod_num_change_cases <- lm(number_change_cases ~ vaccine_rate+home_rate+not_home_rate, data = train)

summary(mod_num_change_cases)
```
```{r data model total deaths}
mod_death <- lm(total_deaths ~ vaccine_rate+home_rate+not_home_rate, data = train)

summary(mod_death)
```
```{r data model num change in deaths}
mod_num_change_deaths <- lm(number_change_death ~ vaccine_rate+home_rate+not_home_rate, data = train)

summary(mod_num_change_deaths)
```

```{r data model pct change in cases}
mod_pct <- lm(pct_change_cases ~ vaccine_rate+home_rate+not_home_rate, data = train)

summary(mod_pct)
```
```{r data model pct change in deaths}
mod_pct_death <- lm(pct_change_death ~ vaccine_rate+home_rate+not_home_rate, data = train)

summary(mod_pct)
```
## Predictions


```{r data model refine pct change in cases}
pred = predict(mod_pct, test, interval = "prediction")
conf = predict(mod_pct, test, interval = "confidence")
pred <- as.data.frame(pred)
mod_pct_prediction <- data.frame(pred,test)
#multiVariateDatawPred <- multiVariateData %>% rowwise() %>% mutate(predicted = #list(predict(mod, interval = "prediction")))

#multiVariateDatawPred$predicted

```

```{r data model refine total deaths}
pred = predict(mod_death, test, interval = "prediction")
conf = predict(mod_death, test, interval = "confidence")
pred <- as.data.frame(pred)
mod_death_prediction <- data.frame(pred,test)
#multiVariateDatawPred <- multiVariateData %>% rowwise() %>% mutate(predicted = #list(predict(mod, interval = "prediction")))

#multiVariateDatawPred$predicted

```

```{r data model refine number change in cases}
pred = predict(mod_num_change_cases, test, interval = "prediction")
conf = predict(mod_num_change_cases, test, interval = "confidence")
pred <- as.data.frame(pred)
mod_num_cases_pred <- data.frame(pred,test)
#multiVariateDatawPred <- multiVariateData %>% rowwise() %>% mutate(predicted = #list(predict(mod, interval = "prediction")))

#multiVariateDatawPred$predicted

```

```{r data model refine number change in deaths}
pred = predict(mod_num_change_deaths, test, interval = "prediction")
conf = predict(mod_death, test, interval = "confidence")
pred <- as.data.frame(pred)
mod_num_death_pred <- data.frame(pred,test)
#multiVariateDatawPred <- multiVariateData %>% rowwise() %>% mutate(predicted = #list(predict(mod, interval = "prediction")))

#multiVariateDatawPred$predicted

```


```{r data model refine pct change in deaths}
pred = predict(mod_pct_death, test, interval = "prediction")
conf = predict(mod_pct_death, test, interval = "confidence")
pred <- as.data.frame(pred)
mod_pct_death_prediction <- data.frame(pred,test)
#multiVariateDatawPred <- multiVariateData %>% rowwise() %>% mutate(predicted = #list(predict(mod, interval = "prediction")))

#multiVariateDatawPred$predicted

```


```{r data model refine total cases}
pred = predict(mod, test, interval = "prediction")
conf = predict(mod, test, interval = "confidence")
pred <- as.data.frame(pred)
total_cases_prediction <- data.frame(pred,test)
#multiVariateDatawPred <- multiVariateData %>% rowwise() %>% mutate(predicted = #list(predict(mod, interval = "prediction")))

#multiVariateDatawPred$predicted

```


## Plot of Predictions vs features


```{r model plot change in number of deaths}

mod_num_death_pred %>% ggplot() + geom_point(aes(x = vaccine_rate, y = number_change_death, color = stateAbbrev))+
  geom_point(aes(x = vaccine_rate, y = fit), color ="purple")

mod_num_death_pred %>% ggplot() + geom_point(aes(x = home_rate, y = number_change_death, color = stateAbbrev))+
  geom_point(aes(x = home_rate, y = fit), color ="purple")

mod_num_death_pred %>% ggplot() + geom_point(aes(x = not_home_rate, y = number_change_death, color = stateAbbrev))+
  geom_point(aes(x = not_home_rate, y = fit), color ="purple")

mod_num_death_pred %>% ggplot() + geom_point(aes(x = stateAbbrev, y = number_change_death, color = stateAbbrev))+
  geom_point(aes(x = stateAbbrev, y = fit), color ="purple")+theme(axis.text.x = element_text(angle = 90))

```
```{r model plot number change in number of cases}

mod_num_cases_pred %>% ggplot() + geom_point(aes(x = vaccine_rate, y = number_change_cases, color = stateAbbrev))+
  geom_point(aes(x = vaccine_rate, y = fit), color ="purple")

mod_num_cases_pred %>% ggplot() + geom_point(aes(x = home_rate, y = number_change_cases, color = stateAbbrev))+
  geom_point(aes(x = home_rate, y = fit), color ="purple")

mod_num_cases_pred %>% ggplot() + geom_point(aes(x = not_home_rate, y = number_change_cases, color = stateAbbrev))+
  geom_point(aes(x = not_home_rate, y = fit), color ="purple")

mod_num_cases_pred %>% ggplot() + geom_point(aes(x = stateAbbrev, y = number_change_cases, color = stateAbbrev))+
  geom_point(aes(x = stateAbbrev, y = fit), color ="purple")+theme(axis.text.x = element_text(angle = 90))

```

```{r model plot pct change in death}

mod_pct_death_prediction %>% ggplot() + geom_point(aes(x = vaccine_rate, y = pct_change_death, color = stateAbbrev))+
  geom_point(aes(x = vaccine_rate, y = fit), color ="purple")

mod_pct_death_prediction %>% ggplot() + geom_point(aes(x = home_rate, y = pct_change_death, color = stateAbbrev))+
  geom_point(aes(x = home_rate, y = fit), color ="purple")

mod_pct_death_prediction %>% ggplot() + geom_point(aes(x = not_home_rate, y = pct_change_death, color = stateAbbrev))+
  geom_point(aes(x = not_home_rate, y = fit), color ="purple")

mod_pct_death_prediction %>% ggplot() + geom_point(aes(x = stateAbbrev, y = pct_change_death, color = stateAbbrev))+
  geom_point(aes(x = stateAbbrev, y = fit), color ="purple")+theme(axis.text.x = element_text(angle = 90))

```

```{r model plot pct change in cases}

mod_pct_prediction %>% ggplot() + geom_point(aes(x = vaccine_rate, y = pct_change_cases, color = stateAbbrev))+
  geom_point(aes(x = vaccine_rate, y = fit), color ="purple")

mod_pct_prediction %>% ggplot() + geom_point(aes(x = home_rate, y = pct_change_cases, color = stateAbbrev))+
  geom_point(aes(x = home_rate, y = fit), color ="purple")

mod_pct_prediction %>% ggplot() + geom_point(aes(x = not_home_rate, y = pct_change_cases, color = stateAbbrev))+
  geom_point(aes(x = not_home_rate, y = fit), color ="purple")

mod_pct_prediction %>% ggplot() + geom_point(aes(x = stateAbbrev, y = pct_change_cases, color = stateAbbrev))+
  geom_point(aes(x = stateAbbrev, y = fit), color ="purple")+theme(axis.text.x = element_text(angle = 90))

```


```{r model plot total cases}

total_cases_prediction %>% ggplot() + geom_point(aes(x = vaccine_rate, y = total_cases, color = stateAbbrev))+
  geom_point(aes(x = vaccine_rate, y = fit), color ="purple")

total_cases_prediction %>% ggplot() + geom_point(aes(x = home_rate, y = total_cases, color = stateAbbrev))+
  geom_point(aes(x = home_rate, y = fit), color ="purple")

total_cases_prediction %>% ggplot() + geom_point(aes(x = not_home_rate, y = total_cases, color = stateAbbrev))+
  geom_point(aes(x = not_home_rate, y = fit), color ="purple")

total_cases_prediction %>% ggplot() + geom_point(aes(x = stateAbbrev, y = total_cases, color = stateAbbrev))+
  geom_point(aes(x = stateAbbrev, y = fit), color ="purple")+theme(axis.text.x = element_text(angle = 90))

plot(total_cases_prediction , pch=20 , cex=1.5 , col="#69b3a2")

```
```{r model plot total deaths}

mod_death_prediction %>% ggplot() + geom_point(aes(x = vaccine_rate, y = total_deaths, color = stateAbbrev))+
  geom_point(aes(x = vaccine_rate, y = fit), color ="purple")

mod_death_prediction %>% ggplot() + geom_point(aes(x = home_rate, y = total_deaths, color = stateAbbrev))+
  geom_point(aes(x = home_rate, y = fit), color ="purple")

mod_death_prediction %>% ggplot() + geom_point(aes(x = not_home_rate, y = total_deaths, color = stateAbbrev))+
  geom_point(aes(x = not_home_rate, y = fit), color ="purple")

mod_death_prediction %>% ggplot() + geom_point(aes(x = stateAbbrev, y = total_deaths, color = stateAbbrev))+
  geom_point(aes(x = stateAbbrev, y = fit), color ="purple")+theme(axis.text.x = element_text(angle = 90))

plot(mod_death_prediction , pch=20 , cex=1.5 , col="#69b3a2")

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
